Working from home and digital divides: resilience in the time of a pandemic

### 2020-12-24 paper/v1_sage/broadband.speed.covid.tex
conclusions

\title{WFH and broadband speed (title needs rework)}

@@

Using the UK as a case study, this paper aims to understand how the
quality and reliability of internet service, as reflected in
\emph{experienced} internet speeds, may reinforce or redress the spatial
and social dimensions of the digital division exposed by the pandemic.
To do so, we employ volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods to
create clusters of UK local authorities with similar temporal signatures
of experienced internet speeds. We then associate these clusters of
local authorities with their socioeconomic and geographic
characteristics to explore how they overlap with or diverge from the
existing economic and digital geography of the UK. Our analysis enables
us to better understand how the spatial and social distribution of both
occupations and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand. We will also
consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era, and broadband services and
infrastructure must be fit for purpose. \textbf{LET'S LEAVE IT FOR NOW,
BUT I THINK WE CAN CRYSTALISE MORE THE RQ}

@@

In this analysis, the terms `telecommuting' and `working from home' are
used interchangeably, as most remote labour during the Covid-19 crisis
was carried out in the homes of individual employees rather than any
other location \citep{eurofound2020}. However, it should be noted that
previous research has explored how telecommuting can occur in other
places, including satellite offices or on public transport
\citep{felstead2012rapid, siha2006telecommuting}. Previous research has
also used a variety of definitions to measure the level of telecommuting
within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or
part-time, and those who use digital technologies to work remotely
full-days or part-days
\citep{allen2015effective, bailey2002review, haddad2009examination}. No
matter the definition, the option and capability to telecommute or work
from home has never been equally distributed spatially or
socio-economically any more than different industries and employment
opportunities have. For example, studies from the United States, the
Netherlands, and the UK found that telecommuters are most likely to hold
professional, managerial, and technical occupations where the workforce
is better educated and wealthier, and that there is suppressed demand
among women and part-time workers
\citep{headicar2016move, peters2004employees, singh2013modeling}.

@@

Based on the outcomes of the time-series clustering, we identify \(J\)
distinct and crisp clusters. We then regress this cluster membership
against a vector \(X_{i}\) of socio-economic and geographic variables,
which are discussed in detail in the relevant Section
{[}4.2{]}{]}(\#sec:4.2), in order to explore how the different patterns
might support or undermine efforts to work from home and maintain safe
productivity and whether they reinforce existing spatial and social
inequalities. This analysis enables us to provide a more nuanced
understanding of how telecommuting and technology intersect at a time of
extreme demand, and what lessons this time has for a future where
telecommuting is likely to remain a common means of accessing work and
broadband services, as well as infrastructure, must be fit for purpose.

@@

The second largest cluster, comprising \(126\) local authorities and
over \(20\) million people, is cluster \(13\). Cluster \(13\) has the
slowest aggregate mean upload speed of any of the clusters, and the
second highest ratio of the standard deviation to the mean. This
suggests that those living in local authorities in this cluster
experienced some of the lowest quality broadband services in terms of
upload speeds in the UK. However, the variation in upload speeds in
cluster \(13\), which can be an indication of its reliability, does not
seem to disproportionately affect the morning peak from
\(9:00\)-\(10:59\), as upload speeds are, on average, only \(2.6\)\%
slower than in the evening peak period between \(19:00\) and \(20:59\),
when entertainment purposes are likely to be using the most bandwidth.
In comparison, the five LADs that are home to \(343\) thousand people in
cluster \(1\) not only experience the second slowest mean upload speeds
and the highest ratio of standard deviation to the mean, but are also
much more affected during the morning peak.

Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
LADs home to \(29.5\) million people, experience aggregate mean upload
speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
lower than the other three large clusters and most of the smaller
clusters, suggesting a middling quality of service. However, the time
profile for cluster \(6\) in Figure \ref{UpClusterL} shows that upload
speeds are highest at \(6:00\) on a Monday morning and peak between
\(23:00\) and midnight on Wednesday and Thursday, but tend to be lower
during the working day. Indeed, experienced mean upload speeds in the
morning peak are \(4.4\)\% lower than in the evening peak -- a greater,
more noticeable change than any of the other large clusters experience,
but smaller than any of the clusters with temporal profiles shown in
Figure \ref{UpClusterS}. Clusters \(8\) and \(12\) also have mean upload
speeds under \(10\)Mb/s, but higher than clusters \(1\) and \(13\), and
their standard deviation is not dissimilar. However, this masks great
variation in when lower speeds are experienced, with the mean upload
speeds much lower between \(9:00-10:59\) than between \(19:00-20:59\) in
cluster \(8\), but slightly faster in the morning in cluster \(12\).

\begin{figure}
\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-5}
\end{figure}

@@

Using auxiliary regressions, we test whether the clusters that have
higher mean speeds and more reliable services consist of LADs that are
more urban, affluent, and / or more likely to benefit from a choice of
high quality internet services. We also estimate which of our clusters
are more likely to have a higher proportion of occupations which enable
telecommuting because of the nature of the work. The results of these
auxiliary regressions are presented in Table \ref{aux}. The dependent
variable is the LAD cluster membership as described in the
\protect\hyperlink{sec:3}{methods and data} section and equation
\ref{eq1} and each column represent a different cluster. The reference
case is cluster \(4\), which includes only the local authority of
Hambleton in North Yorkshire, a rural area of just over ninety thousand
people. Mean, experienced upload speeds in cluster \(4\) are close to
both the average speeds for the \(13\) clusters (\(9.9\)Mb/s) and the
pre-clustered average for the whole sample (\(9.3\)Mb/s) as shown in
Table \ref{up.cluster.descr}. However, the standard deviation for
cluster \(4\) and the difference between average speeds in the morning
compared to the evening peak periods are indications of worse
reliability than many of the other clusters. Hence, the results in Table
\ref{aux} should be seen as relative rather than absolute probabilities.

First, the number of speed tests run per cluster inhabitant between
\(9:00\)-\(10:59\) is an indication of satisfaction or at least a lack
of concern over broadband speeds and quality of service. People in LADs
assigned to clusters \(6\) and \(13\) were more likely to run speed
tests (at a per capita basis) at this time of day, whilst people in
cluster \(11\) were the least likely to do so. This high level of
satisfaction matches the high quality and reliability of service we
would expect from the analysis in Section {[}4.1{]}{]}(\#sec:4.1). Those
in clusters \(2\), \(3\), \(9\) and \(10\) were also less likely to ran
speed tests per capita than those in the other six multi-authority
clusters. Cluster \(7\) was the only cluster with higher mean speeds and
middle to high measures of reliability which was more likely to run more
speed tests per capita. This might be a reflection of the choice of ISP
and connection options in those LADs, as, along with cluster \(12\),
those in cluster \(7\) were least likely to have Virgin Media
connections. Fast Virgin Media cable connections have historically only
been available to \(45\)\% of premises in the UK \citep{ofcom2016},
where the more lucrative and competitive market originally attracted the
cable TV provider. Those in clusters \(2\), \(3\), \(9\) and \(11\)
benefit from a higher proportion of Virgin connections, which is an
indication that people in these clusters are more likely to live in
urban areas, with more choice of broadband services, or in other words,
are more likely to be on the right side of the first level,
infrastructure-based digital divide.

We employ distance from London and for the nearest metropolitan area
(including London) as two more variables depicting peripherality, urban
structure and, potentially, first level digital divides. The broadband
speed tests run in the authorities in cluster \(3\) are more likely to
be taking place close to London than those run in any of the other
clusters, a result that makes sense considering two of the four
authorities in cluster \(3\) are the London commuter towns of Harlow and
Luton. However, even though London was also included in the variable
calculating distance from the centre of one of either the ten largest
metropolitan areas in England, or Glasgow or Cardiff, tests run in
cluster \(3\) are likely to be furthest away. Still, whilst significant,
the effect is tiny and the other members of cluster \(3\) are Corby and
Eastbourne, two towns which, whilst not part of metropolitan areas, are
also not in remote parts of the country, and are home to an estimated
population of \(175\) thousand people between them. \textbf{move maps to
here?}

Meanwhile, LADs in cluster \(7\) are most likely to be near the centre
of a large metropolitan area, even though the four local authorities of
cluster \(7\) include no central urban boroughs and only one LAD that is
part of a metropolitan area of governance - Tameside in Greater
Manchester. This may be why those in cluster \(7\) are not served by
Virgin Media. It also is a demonstration of the complexity of
experienced broadband upload speeds as captured by time-series
clustering, and their likely interaction with first level digital
divides. Cluster \(1\), which our analysis suggests lacks broadband
resilience, contains five, mainly rural authorities. However, they are
scattered around the country -- Ceredigion in West Wales, Darlington in
the Northeast, Eden between the Lake District and the North Penines,
Rossendale in the Northwest, and Rutland in the East Midlands -- and
therefore, the results in Table \ref{aux} indicate that these
authorities are not as far from the centre of a metropolitan area as
authorities in cluster \(3\), but furthest from London.

@@

Yet we need to consider the results for other variables in Table
\ref{aux} to better determine whether clusters \(1\) and \(13\), which
appear to suffer most from a lack of quality internet services, are also
more likely to have a low skilled workforce, less able to benefit from
telecommuting. Cluster \(1\) has the lowest propotion of educated people
of any of the clusters, and the lowest earnings, despite recording the
highest proportion of managerial, professional jobs, and the second
highest proportion of tech jobs. Rural areas such as those in cluster
\(1\) are often home to many retired people, which might explain these
results or perhaps, as these figures are relative, we could note that
cluster \(1\) also has a greater proportion of skilled trades than other
clusters. Either way, it seems likely that the first level digital
divide reinforces other inequities in cluster \(1\). In comparison,
those cluster \(13\) are more likely to earn more and have a better
education. Cluster \(13\) also has the most businesses per inhabitant,
but is somewhere in the middle in terms of job density. If this is an
indication of a high number of SMEs, it might give some explanation for
the variable internet quality, considering that small businesses have
not been seen as the most valuable customers for higher speed broadband
services \textbf{insert ref}.

Cluster \(6\) is the largest cluster, and thus, like cluster \(13\),
there is more noise within the averages we use to measure internet
quality and reliability, and the result indicates an average level of
service, even if better than LADs in cluster \(13\). Cluster \(6\) also
falls towards the middle of the clusters on many of the socio- economic
variables in Table \ref{aux}. It ranks third or fourth out of the eleven
clusters in the table in terms of having a higher proportion of
managerial, professional, and tech jobs, as well as higher earnings, but
is fourth from bottom in terms of education. The LADs in cluster \(6\)
are also diverse, with few truly remote areas, but many urban areas in
both the North, including Birmingham, Leeds and Sheffield, and South,
including twelve London Boroughs, and many suburban areas and smaller
cities like Oxford and Cambridge. The capital cities of the other UK
nations, Belfast, Cardiff and Edinburgh, are also in this cluster,
suggesting perhaps that the lower level of reliability discussed in
Section \protect\hyperlink{sec:4.1}{4.1} may be as a result of increased
demand, e.g.~for telecommuting.

In comparison, clusters \(9\) and \(11\), which benefit from the most
resilient internet connections, are characterised by lower share of
occupations that are\\
may benefit from the ability to telecommute (e.g.~managerial and
technical). Cluster \(9\) is more highly educated and higher earning
than cluster \(11\), has fewer businesses per inhabitant and less job
density. Indeed, cluster \(11\) has a negative coefficient for the NVQ4
variable, whilst for cluster \(9\) the coefficient is positive, and
cluster \(11\) has the second highest job density and cluster \(9\) the
second lowest. The former might be an indication of the higher returns
that resilient broadband infrastructure generates for cluster \(9\)
versus cluster \(11\). Indeed, we did detect slightly more slowdown in
the AM peak in cluster \(9\) than in cluster \(11\) (see Table
\ref{up.cluster.descr}). Meanwhile, a propensity for a high job density
might indicate more labour intensive economic structures in cluster
\(11\) than in the LADs in cluster \(9\).

The LADs in clusters \(9\) and \(11\) also have opposite signs for their
likelihood of being in the South of England. Those in cluster \(9\) less
likely to be in the South and those in cluster \(11\) more likely,
although the coefficient for cluster \(9\) might be skewed by the
presence of the Scottish cities of Glasgow and Aberdeen, as Scotland has
a different economic profile than England. Still, both clusters consist
of larger LADs in terms of population, including the urban centre of
Manchester and three of the nine other districts of Greater Manchester,
Nottingham, both Portsmouth and Southampton, eight London Boroughs, and
four of the seven constituent authorities of the West Midlands Combined
Authority. Among the \(65\) LADs are also a number of other tightly
bounded urban areas, such as Blackpool, Ipswich, Norwich, Slough, and
Stevenage, and urban areas at the centre of less confined districts like
Burnley, the digital hub of Milton Keynes, and Northampton. Thus, our
findings confirm the importance of geography and urban locations for
being on the right side of the first level digital divide, and suggest
that internet resilience cuts across other digital and economic divides.

@@

In comparison, LADs in cluster \(12\) are less likely to be able to
benefit from quality internet services if they had them, with fewer
individuals achieving NVQ4 or better and lower levels of occupations
that would benefit from homeworking. Yet LADs in this cluster benefit
from the second highest level of earnings, and have the largest stock of
businesses per inhabitant and the highest job density. This density of
businesses could be why cluster \(12\) is the only cluster with higher
speeds in the morning peak during the study period. If people are at
home, business premises might well be abandoned. Cluster \(12\) is home
to \(1.5\)million people spread across twelve LADs, including the City
of Westminster and the London Borough of Hammersmith. However, it also
includes a LAD in Northern Ireland, two in Wales, two suburban areas
North of London, Preston in the Northwest, East Staffordshire in the
Midlands, and Fenland in the East of England. This spatial diversity
demonstrates that the temporal clustering of internet resilience does
not create spatial clusters, and digital divides do not necessarily
overlap with economic ones.

@@

Upload speeds are not used as benchmarks in the UK, nor have been seen
as integral to universal service, considering there has never before
been such extreme demand for telecommuting and operations such as video
calls.

@@

The long-term effects of such drastic changes in telecommuting and
attitudes towards working from home are difficult to predict.
Nevertheless, they span through various aspects of economy and society:
from changes to transportation planning due to altered commuting
patters, to changes in land use and urban planning to accommodate people
who work from home \citep{BUDNITZ2020102713}\textbf{also 2020 Swedish
article from JTG}; and from productivity and innovation changes, to
changes in agglomeration externalities and the attraction of large
cities \citep{econobs} just to name a few. This paper is positioned to
support endeavours in understanding the effects of increased
telecommuting by exposing the spatial and social dimensions of
telecommuting including the resilience of broadband speeds in terms of
both quality and reliability of service, and whether this reinforces or
redresses prior digital divisions. \textbf{took this bit you wrote to
put in the discussion at the end?}

>>>>

\title{Working from home and digital divides: resilience in the time of a
pandemic}

@@

Using the UK as a case study, this paper aims to understand how the
quality and reliability of internet service, as reflected in
\emph{experienced} internet speeds, may reinforce or redress the spatial
and social dimensions of the digital division exposed by the pandemic.
To do so, we employ volunteered geographic data on individual broadband
speed tests and state-of-the-art time-series clustering methods to
create clusters of UK local authorities with similar temporal signatures
of experienced internet speeds. We then associate these clusters of
local authorities with their socioeconomic and geographic
characteristics to explore how they overlap with or diverge from the
existing economic and digital geography of the UK. Our analysis enables
us to better understand how the spatial and social distribution of both
occupations and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand. We will also
consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era, and broadband services and
infrastructure must be fit for purpose.

@@

In this analysis, the terms `telecommuting' and `working from home\^{}
{[}See also the popular social media hashtag \#WFM{]}' are used
interchangeably, as most remote labour during the Covid-19 crisis was
carried out in the homes of individual employees rather than any other
location \citep{eurofound2020}. However, it should be noted that
previous research has explored how telecommuting can occur in other
places, including satellite offices or on public transport
\citep{felstead2012rapid, siha2006telecommuting}. Previous research has
also used a variety of definitions to measure the level of telecommuting
within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or
part-time, and those who use digital technologies to work remotely
full-days or part-days
\citep{allen2015effective, bailey2002review, haddad2009examination}. No
matter the definition, the option and capability to telecommute or work
from home has never been equally distributed spatially or
socio-economically any more than different industries and employment
opportunities have. For example, studies from the United States, the
Netherlands, and the UK found that telecommuters are most likely to hold
professional, managerial, and technical occupations where the workforce
is better educated and wealthier, and that there is suppressed demand
among women and part-time workers
\citep{headicar2016move, peters2004employees, singh2013modeling}.

@@

Based on the outcomes of the time-series clustering, we identify \(J\)
distinct and crisp clusters. We then regress this cluster membership
against a vector \(X_{i}\) of socio-economic and geographic variables,
which are discussed in detail in the relevant Section
\protect\hyperlink{sec:4.2}{4.2}, in order to explore how the different
patterns might support or undermine efforts to work from home and
maintain safe productivity and whether they reinforce existing spatial
and social inequalities. This analysis enables us to provide a more
nuanced understanding of how telecommuting and technology intersect at a
time of extreme demand, and what lessons this time has for a future
where telecommuting is likely to remain a common means of accessing work
and broadband services, as well as infrastructure, must be fit for
purpose.

Our analysis is not free of limitations as, for instance, although we do
have individual data we cannot identify individuals or households and,
therefore, we need to aggregate our data at the LAD level. However, such
individual data could only be found though surveys which by definition
result in limited samples and, importantly, less detailed data about the
quality of internet connectivity.

@@

The second largest cluster, comprising \(126\) local authorities and
over \(20\) million people, is cluster \(13\), which has the slowest
aggregate mean upload speed of any of the clusters, and the second
highest ratio of the standard deviation to the mean. This suggests that
those living in local authorities in this cluster experienced some of
the lowest quality broadband services in terms of upload speeds in the
UK. However, the variation in upload speeds in cluster \(13\), which can
be an indication of its reliability, does not seem to disproportionately
affect the morning peak from \(9:00\)-\(10:59\), as upload speeds are,
on average, only \(2.6\)\% slower than in the evening peak period
between \(19:00\) and \(20:59\), when entertainment purposes are likely
to be using the most bandwidth. In comparison, the five LADs that are
home to \(343\) thousand people in cluster \(1\) not only experience the
second slowest mean upload speeds and the highest ratio of standard
deviation to the mean, but are also much more affected during the
morning peak.

Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
LADs home to \(29.5\) million people, experienced aggregate mean upload
speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
lower than the other three large clusters and most of the smaller
clusters, suggesting a middling quality of service. However, the
temporal profile for cluster \(6\) in Figure \ref{UpClusterL} shows that
upload speeds are highest at \(6:00\) on a Monday morning and peak
between \(23:00\) and midnight on Wednesday and Thursday, but tend to be
lower during the working day. Indeed, experienced mean upload speeds in
the morning peak are \(4.4\)\% lower than in the evening peak -- a
greater, more noticeable change than any of the other large clusters
experience, but smaller than any of the clusters with temporal profiles
shown in Figure \ref{UpClusterS}. Clusters \(8\) and \(12\) also have
mean upload speeds under \(10\)Mb/s, but higher than clusters \(1\) and
\(13\), and their standard deviation is not dissimilar. However, this
masks great variation in when lower speeds are experienced, with the
mean upload speeds much lower between \(9:00-10:59\) than between
\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning in
cluster \(12\).

@@

Using an auxiliary multinomial regression, we test whether the clusters
that have higher mean speeds and more reliable services consist of LADs
that are more urban, affluent, and / or more likely to benefit from a
choice of high quality internet services. We also estimate which of our
clusters are more likely to have a higher proportion of occupations
which enable telecommuting because of the nature of the work. The
results of the auxiliary regression are presented in Table \ref{aux}.
The dependent variable is the LAD cluster membership as described in the
\protect\hyperlink{sec:3}{methods and data} section and equation
\ref{eq1} and each column represent a different cluster. The reference
case is cluster \(4\), which includes only the local authority of
Hambleton in North Yorkshire, a rural area of just over ninety thousand
people. Mean, experienced upload speeds in cluster \(4\) are close to
both the average speeds for the \(13\) clusters (\(9.9\)Mb/s) and the
pre-clustered average for the whole sample (\(9.3\)Mb/s) as shown in
Table \ref{up.cluster.descr}. However, the standard deviation for
cluster \(4\) and the difference between average speeds in the morning
compared to the evening peak periods are indications of worse
reliability than many of the other clusters. Hence, the results in Table
\ref{aux} should be seen as relative rather than absolute probabilities.

First, we control for the number of speed tests run per cluster
inhabitant between \(9:00\)-\(10:59\) as well as the share of fast
Virgin Media internet connections. Regarding the former, we expect
people in LADs with more unreliable connections to test more often their
internet speeds and the results validate our priors. Regarding the
former, fast Virgin Media cable connections have historically only been
available to \(45\)\% of premises in the UK \citep{ofcom2016}, where the
more lucrative and competitive market originally attracted the cable TV
provider. Those in clusters \(2\), \(3\), \(9\) and \(11\) benefit from
a higher proportion of Virgin connections, which is an indication that
people in these clusters are more likely to live in urban areas, with
more choice of broadband services, or in other words, are more likely to
be on the right side of the first level, infrastructure-based digital
divide.

Then, we employ distance from London and for the nearest metropolitan
area (including London) as two variables depicting peripherality, urban
structure and, potentially, first level digital divides. The broadband
speed tests run in the authorities in cluster \(3\) are more likely to
be taking place close to London than those run in any of the other
clusters, a result that makes sense considering two of the four
authorities in cluster \(3\) are the London commuter towns of Harlow and
Luton. However, even though London was also included in the variable
calculating distance from the centre of one of either the ten largest
metropolitan areas in England, or Glasgow or Cardiff, tests run in
cluster \(3\) are likely to be furthest away. Still, whilst significant,
the effect is tiny and the other members of cluster \(3\) are Corby and
Eastbourne, two towns which, whilst not part of metropolitan areas, are
also not in remote parts of the country, and are home to an estimated
population of \(175\) thousand people between them.

\begin{figure}
\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-6}
\end{figure}

Meanwhile, LADs in cluster \(7\) are most likely to be near the centre
of a large metropolitan area, even though the four local authorities of
cluster \(7\) include no central urban boroughs and only one LAD that is
part of a metropolitan area of governance - Tameside in Greater
Manchester. This may explain why those in cluster \(7\) are not served
by Virgin Media. It also is a demonstration of the complexity of
experienced broadband upload speeds as captured by time-series
clustering, and their likely interaction with first level digital
divides. Cluster \(1\), which our analysis suggests lacks broadband
resilience, contains five, mainly rural authorities. However, they are
scattered around the country -- Ceredigion in West Wales, Darlington in
the Northeast, Eden between the Lake District and the North Penines,
Rossendale in the Northwest, and Rutland in the East Midlands -- and
therefore, the results in Table \ref{aux} indicate that these
authorities are not as far from the centre of a metropolitan area as
authorities in cluster \(3\), but furthest from London.

@@

Yet we need to consider the results for other variables in Table
\ref{aux} to better determine whether clusters \(1\) and \(13\), which
appear to suffer most from a lack of quality internet services, are also
more likely to have a low skilled workforce, less able to benefit from
telecommuting. Cluster \(1\) has the lowest propotion of educated people
of any of the clusters, and the lowest earnings, despite recording the
highest proportion of managerial, professional jobs, and the second
highest proportion of tech jobs. Rural areas such as those in cluster
\(1\) are often home to many retired people, which might explain these
results or perhaps, as these figures are relative, we could note that
cluster \(1\) also has a greater proportion of skilled trades than other
clusters. Either way, it seems likely that the first level digital
divide reinforces other inequities in cluster \(1\). In comparison,
those in cluster \(13\) are more likely to earn more and have a better
education. Cluster \(13\) also has the most businesses per inhabitant,
but is somewhere in the middle in terms of job density. If this is an
indication of a high number of SMEs, it can explain the variable
internet quality, considering that small businesses have not been seen
as the most valuable customers for higher speed broadband services
\textbf{insert ref}.

Cluster \(6\) is the largest cluster, and thus, like cluster \(13\),
there is more noise within the averages we use to measure internet
quality and reliability, and the result indicates an average level of
service, even if better than LADs in cluster \(13\). Cluster \(6\) also
falls towards the middle of the clusters on many of the socio- economic
variables in Table \ref{aux}. It ranks third or fourth out of the eleven
clusters in the table in terms of the likelihood of having a higher
proportion of managerial, professional, and tech jobs, as well as higher
earnings, but is fourth from bottom in terms of education. The LADs in
cluster \(6\) are also diverse, with few truly remote areas, but many
urban areas in both the North, including Birmingham, Leeds and
Sheffield, and South, including twelve London Boroughs, and many
suburban areas and smaller cities like Oxford and Cambridge. The capital
cities of the other UK nations, Belfast, Cardiff and Edinburgh, are also
in this cluster, suggesting perhaps that the lower level of reliability
discussed in Section \protect\hyperlink{sec:4.1}{4.1} may be as a result
of increased demand, e.g.~for telecommuting.

In comparison, clusters \(9\) and \(11\), which enjoy the most resilient
internet connections, are characterised by higher share of occupations
that may benefit from the ability to telecommute (e.g.~managerial).
Also, LADs in Cluster \(9\) are more likely to host highly educated
individuals with higher earning than cluster \(11\), which has fewer
businesses per inhabitant and lower job density. Indeed, cluster \(11\)
has a negative coefficient for the NVQ4 variable, whilst for cluster
\(9\) the coefficient is positive, and cluster \(11\) has the second
highest job density and cluster \(9\) the second lowest. The former
might be an indication of the higher returns that resilient broadband
infrastructure can generate for cluster \(9\) versus cluster \(11\).
Indeed, we did detect slightly more slowdown in the AM peak in cluster
\(9\) than in cluster \(11\) (see Table \ref{up.cluster.descr}).
Meanwhile, a propensity for a higher job density indicate higher
urbanization rate in LADs in cluster \(11\) than those in cluster \(9\).

LADs in cluster \(9\) are also less likely to be in the South and those
in cluster \(11\) more likely, although the coefficient for cluster
\(9\) might be skewed by the presence of the Scottish cities of Glasgow
and Aberdeen, as Scotland has a different economic profile than England.
Still, both clusters consist of larger LADs in terms of population,
including the urban centre of Manchester and three of the nine other
districts of Greater Manchester, Nottingham, both Portsmouth and
Southampton, eight London Boroughs, and four of the seven constituent
authorities of the West Midlands Combined Authority. Among the \(65\)
LADs are also a number of other tightly bounded urban areas, such as
Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas at
the centre of less confined districts like Burnley, the digital hub of
Milton Keynes, and Northampton. Thus, our findings confirm the
importance of geography and urban locations for being on the right side
of the first level digital divide, and suggest that internet resilience
cuts across other digital and economic divides.

@@

In comparison, LADs in cluster \(12\) are less likely to be able to
benefit from quality internet services if they had them, with fewer
individuals achieving NVQ4 or better and lower levels of occupations
that would benefit from homeworking. Yet LADs in this cluster benefit
from the second highest level of earnings, and have the largest stock of
businesses per inhabitant and the highest job density. This density of
businesses could be why cluster \(12\) is the only cluster with higher
speeds in the morning peak during the study period. If people are at
home, business premises might well be abandoned. Cluster \(12\) is home
to \(1.5\)million people spread across twelve LADs, including the City
of Westminster and the London Borough of Hammersmith. However, it also
includes a LAD in Northern Ireland, two in Wales, two suburban areas
North of London, Preston in the Northwest, East Staffordshire in the
Midlands, and Fenland in the East of England. This spatial diversity
demonstrates that the temporal clustering of internet resilience is not
necessarily spatially dependent, and digital divides do not necessarily
overlap with economic ones.

@@

This paper offers a new perspective on telecommuting from the viewpoint
of the complex web of digital divides. Good and reliable internet
connections are necessary for the population to be able to work from
home, but at the same time not every place hosts individuals in
occupations which allow for telecommuting and not every place hosts
individuals with the necessary skills to effectively use the internet
and work form home. Importantly, not every place can utilise good
internet connectivity -- even if they have them -- in order to achieve
economic resilience in a period like the current pandemic, but also for
a future when working from home might be a more common activity.

Contrary to the previous literature, we employ experienced upload speeds
as a benchmark of internet resilience concerning the capacity to work
from home. Upload speeds are not used as benchmarks in the UK, nor have
been seen as integral to universal service, considering there has never
before been such extreme demand for telecommuting and operations such as
video calls. While download speeds are associated more with video
streaming services, fast upload internet speed are necessary for
seamless video calls and uploading large files, activities associated
with work. Given the granular nature of our data, we are able to
consider upload internet speeds during working hours. To analyse these
data we employ time series clustering methods, a family of unsupervised
machine learning techniques which are not often utilised in geographical
research.

Using the UK as a case study, our analysis illustrated a nuanced
picture, which is the outcome of the intersection between the different
digital divides and the potential to telecommute. While some LADs appear
to benefit both from reliable internet connections and population which
can work from home and, in essence, capitalise the digital
infrastructure, other LADs are suffering by both unreliable internet and
low share of individuals who can potentially telecommute.

\begin{itemize}
\item
  different places have different levels of preparedness, some places
  have some ground to cover
\item
  benefits
\end{itemize}

@@

The long-term effects of such drastic changes in telecommuting and
attitudes towards working from home are difficult to predict.
Nevertheless, they span through various aspects of economy and society:
from changes to transportation planning due to altered commuting
patters, to changes in land use and urban planning to accommodate people
who work from home \citep{BUDNITZ2020102713, ELLDER2020102777} and from
productivity and innovation changes, to changes in agglomeration
externalities and the attraction of large cities \citep{econobs} just to
name a few. The take home message from this paper is that working form
home and, consequently, its long term effects cannot be understood, set
aside plan our cities for, without considering the underpinning digital
divides and how telecommuting intersects with them.
